---
title: "How concentrated is St. Petersburg's mapped park area?"
output: html_document
---

The checked `stpete-parks` layer has park polygons, including preserves and a golf course. This story asks what share of the **sum of mapped polygon area** comes from the ten largest features. It uses a local projected CRS for area, retrieves every matching row, and checks the package's completeness record before drawing a chart.

**Source:** [City of St. Petersburg Parks layer](https://services2.arcgis.com/9qPLjNtocjo438CJ/arcgis/rest/services/Parks_view/FeatureServer/0) and [publisher item](https://csp.maps.arcgis.com/home/item.html?id=ff0015e3a49044ecbbd7a758e8098076). Knit with the `tampaBayOpenData` and `sf` packages installed. The live source can change between runs.

```{r s02-step-1}
library(tampaBayOpenData)
library(sf)

info <- tbod_dataset_info("stpete-parks")
cat(info$publisher, "|", info$title, "\nSource:", info$source_url, "\n")
```

## Retrieve the complete layer

Request geometry in EPSG:26917 (UTM zone 17N), whose metre coordinates support local area calculation. The checked entry verifies the upstream schema, and full retrieval validates the object-ID manifest.

```{r s02-step-2}
parks <- tbod_get_dataset(
  "stpete-parks",
  fields = c("OBJECTID", "NAME", "FULLADDR"),
  spatial = TRUE, out_sr = 26917, limit = Inf
)
p <- tbod_provenance(parks)
stopifnot(isTRUE(p$complete), p$matched_rows == nrow(parks),
          p$returned_rows == nrow(parks), sf::st_crs(parks)$epsg == 26917)
print(data.frame(retrieved_at_utc = format(p$retrieved_at, tz = "UTC"),
                 matched_rows = p$matched_rows, returned_rows = p$returned_rows,
                 complete = p$complete, integrity = p$integrity))
```

## Rank polygon footprints

Area is calculated from the returned geometries, not the service's stored area field. A row is a mapped polygon; this is not a measure of public access, facilities, or area per resident.

```{r s02-step-3}
area_ha <- as.numeric(sf::st_area(parks)) / 10000
stopifnot(all(is.finite(area_ha)), all(area_ha > 0))
ranked <- data.frame(name = parks$NAME, area_ha = area_ha)
ranked <- ranked[order(ranked$area_ha, decreasing = TRUE), ]
print(data.frame(polygons = nrow(ranked), unique_names = length(unique(ranked$name)),
                 median_hectares = round(median(ranked$area_ha), 2),
                 top_ten_area_share = round(sum(head(ranked$area_ha, 10)) /
                                              sum(ranked$area_ha), 3)))
print(transform(head(ranked, 10), area_ha = round(area_ha, 2)), row.names = FALSE)
```

```{r s02-step-4, fig.width=8, fig.height=5}
top_ten <- head(ranked, 10)
old_par <- par(mar = c(4.5, 16, 3, 1))
barplot(rev(top_ten$area_ha), names.arg = rev(top_ten$name), horiz = TRUE,
        las = 1, cex.names = 0.9, col = "#267B82", border = NA,
        xlab = "Mapped polygon area (hectares)",
        main = "Ten largest mapped park polygons")
par(old_par)
```

The largest mapped features include nature preserves and a golf course. The sum can double-count overlap if polygons overlap; it is not a dissolved land-cover estimate. A new run should report the current ten features and share instead of assuming the saved ranking still holds.

**Observed on October 6, 2026:** The complete pull returned 182 park polygons with 182 distinct names. The ten largest polygons accounted for 49.6% of the sum of mapped polygon areas. Their ranking may change when the source changes.
