---
title: "A paired look at Hillsborough water-quality readings"
output: html_document
---

The Hillsborough County Environmental Protection Commission publishes its latest staff-reviewed monitoring samples. This report asks whether the two reported dissolved-oxygen fields, DO_T and DO_B, differ within records where both are present. The layer is a current sample view, so these records cannot support a time trend.

Source: [Hillsborough EPC water-quality layer](https://services.arcgis.com/apTfC6SUmnNfnxuF/arcgis/rest/services/EPC_WQM_DATA/FeatureServer/0). The package's checked catalog supplies the publisher attribution and scope note.

## 1. Inspect the checked source

The bundled schema check catches a missing or renamed field before we make the live request. Field labels are kept exactly as published; their units and sampling methods need the publisher's documentation.

```{r s04-step-1}
library(tampaBayOpenData)

id <- "hillsborough-water-quality"
fields <- c("StationNumber", "AreaName", "DO_T", "DO_B")
info <- tbod_dataset_info(id)
schema <- tbod_schema(id)
stopifnot(all(fields %in% schema$fields$name))
cat(info$publisher, "-", info$title, "\n")
cat("Source:", info$source_url, "\n")
cat("Scope:", info$scope_note, "\n")
```

## 2. Retrieve the full matching layer

A full pull lets us compare every returned row with the package's object-ID manifest. The result is live and may change on a later run.

```{r s04-step-2}
samples <- tbod_get_dataset(id, fields = fields, total_timeout = 180)
source <- tbod_provenance(samples)
stopifnot(isTRUE(source$complete),
          source$returned_rows == nrow(samples),
          source$matched_rows == nrow(samples))
cat(sprintf("%s records retrieved at %s; integrity: %s\n",
            format(nrow(samples), big.mark = ","),
            format(source$retrieved_at, tz = "UTC", usetz = TRUE),
            source$integrity))
```

## 3. Count usable pairs

Missing values matter here: a comparison between DO_T and DO_B is only paired when both values were reported on the same row. We show the exclusions before plotting.

```{r s04-step-3}
paired <- samples[is.finite(samples$DO_T) & is.finite(samples$DO_B), , drop = FALSE]
station <- trimws(as.character(samples$StationNumber))
station <- station[!is.na(station) & nzchar(station)]
checks <- data.frame(
  measure = c("All records", "Distinct nonblank station IDs",
              "Missing DO_T", "Missing DO_B", "Usable pairs"),
  count = c(nrow(samples), length(unique(station)),
            sum(!is.finite(samples$DO_T)), sum(!is.finite(samples$DO_B)),
            nrow(paired))
)
print(checks, row.names = FALSE)
if (nrow(paired)) {
  difference <- paired$DO_T - paired$DO_B
  cat(sprintf("Median DO_T - DO_B among usable pairs: %.2f\n",
              median(difference)))
}
```

## 4. Compare the two reported fields

The diagonal marks equal reported values. This is an exploratory field comparison, not a water-quality classification or a test of improvement.

```{r s04-step-4, fig.width=7, fig.height=4}
if (nrow(paired)) {
  limits <- range(c(paired$DO_T, paired$DO_B))
  plot(paired$DO_B, paired$DO_T, xlim = limits, ylim = limits,
       pch = 16, col = grDevices::rgb(0.11, 0.39, 0.57, 0.45),
       xlab = "Reported DO_B", ylab = "Reported DO_T",
       main = "Paired dissolved-oxygen fields")
  abline(0, 1, lty = 2, col = "gray40")
  cat(sprintf("%d of %d usable pairs have DO_T above DO_B.\n",
              sum(paired$DO_T > paired$DO_B), nrow(paired)))
} else {
  plot.new()
  text(0.5, 0.5, "No records have both oxygen fields")
}
```

## What this shows

The chart describes only rows with both measurements; the count table shows how much of the layer that covers. The source describes the **latest reviewed samples**, not a full station history. Units, detection limits, methods, and uneven station coverage need source review before environmental comparisons or policy use.

**Observed on October 6, 2026:** The complete pull returned 263 records. Of these, 161 had both reported oxygen values; the median `DO_T - DO_B` in those pairs was 0.26. Source methods and units need review before interpreting that difference.
