---
title: "Assisted units in Pinellas housing records"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error = FALSE)
```

The Pinellas County assisted-housing layer reports total units and assisted units for mapped housing records. This report asks how often both counts are usable and how their reported values compare. It treats each polygon row as a mapped record, because repeated site identifiers could otherwise double-count units.

Source: [Pinellas assisted-housing layer](https://services.arcgis.com/f5HgUpxURgEzTccH/ArcGIS/rest/services/Pinellas_AssistedHousing_view/FeatureServer/0). The package's checked catalog provides the publisher attribution and scope note.

## 1. Inspect the checked source

The numeric unit fields and source identifiers are verified against the bundled schema. Their values come from the live publisher service when the next cell runs.

```{r inspect_source}
library(tampaBayOpenData)

id <- "pinellas-assisted-housing"
fields <- c("PID", "DEVELOPMENT", "CITY", "HOUSINGPROGRAM",
            "TOTALUNITS", "ASSISTEDUNITS")
info <- tbod_dataset_info(id)
schema <- tbod_schema(id)
stopifnot(all(fields %in% schema$fields$name))
cat(info$publisher, "-", info$title, "\n")
cat("Source:", info$source_url, "\n")
cat("Scope:", info$scope_note, "\n")
```

## 2. Retrieve the full matching layer

The provenance check requires a complete result before interpreting its distribution. It still represents a live service rather than a frozen historical snapshot.

```{r retrieve_housing}
homes <- tbod_get_dataset(id, fields = fields, total_timeout = 180)
source <- tbod_provenance(homes)
stopifnot(isTRUE(source$complete),
          source$returned_rows == nrow(homes),
          source$matched_rows == nrow(homes))
cat(sprintf("%s mapped records retrieved at %s; integrity: %s\n",
            format(nrow(homes), big.mark = ","),
            format(source$retrieved_at, tz = "UTC", usetz = TRUE),
            source$integrity))
```

## 3. Check the unit fields

A share is defined only when total units are positive and assisted units fall between zero and total units. Missing and inconsistent rows remain visible in the audit instead of silently disappearing.

```{r audit_unit_fields}
total <- homes$TOTALUNITS
assisted <- homes$ASSISTEDUNITS
both <- is.finite(total) & is.finite(assisted)
valid <- both & total > 0 & assisted >= 0 & assisted <= total
pid <- trimws(as.character(homes$PID))
pid <- pid[!is.na(pid) & nzchar(pid)]
checks <- data.frame(
  measure = c("All mapped records", "Missing either unit count",
              "Zero total units with both counts", "Inconsistent unit counts",
              "Usable unit pairs", "Repeated nonblank PID values"),
  count = c(nrow(homes), sum(!both), sum(both & total == 0),
            sum(both & (total < 0 | assisted < 0 | assisted > total)),
            sum(valid), sum(duplicated(pid)))
)
print(checks, row.names = FALSE)
usable <- homes[valid, , drop = FALSE]
if (nrow(usable)) {
  share <- usable$ASSISTEDUNITS / usable$TOTALUNITS
  cat(sprintf("Median assisted share among usable mapped records: %.1f%%\n",
              100 * median(share)))
}
```

## 4. Compare counts within each mapped record

The diagonal represents equal assisted and total units. Points below it report fewer assisted units than total units. No unit counts are summed across polygons, so repeated site identifiers cannot inflate a regional total here.

```{r plot_unit_counts, fig.width=7, fig.height=5}
if (nrow(usable)) {
  upper <- max(c(usable$TOTALUNITS, usable$ASSISTEDUNITS))
  plot(usable$TOTALUNITS, usable$ASSISTEDUNITS,
       xlim = c(0, upper), ylim = c(0, upper),
       pch = 16, col = grDevices::rgb(0.55, 0.27, 0.28, 0.5),
       xlab = "Reported total units", ylab = "Reported assisted units",
       main = "Unit counts in mapped housing records")
  abline(0, 1, lty = 2, col = "gray40")
  cat(sprintf("%d of %d usable records report fewer assisted than total units.\n",
              sum(usable$ASSISTEDUNITS < usable$TOTALUNITS), nrow(usable)))
} else {
  plot.new()
  text(0.5, 0.5, "No records have usable unit pairs")
}
```

## What this shows

The analysis describes the publisher's mapped housing records, with each polygon row shown once. Program participation and unit counts can change. The layer does not establish current eligibility, vacancies, or a complete affordable-housing inventory. Repeated PIDs in the audit warrant site-level review before aggregating unit totals.

**Observed on 2026-10-06:** Of 191 mapped records, 151 had usable unit pairs and 40 were missing at least one count. Seven nonblank PID occurrences repeated an earlier value; 33 of the 151 usable records reported fewer assisted than total units. Rerunning this report will calculate the current figures.
