---
title: "What kinds of records are in the regional data-center inventory?"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error = FALSE)
```

The [Tampa Bay Regional Planning Council layer](https://services3.arcgis.com/RgIBbQIF0Y6T3AX6/arcgis/rest/services/Tampa_Bay_Region_Data_Centers/FeatureServer/0) mixes locations and project statuses. This story checks status by county and audits `Estimated_Size` before any size comparison. It is a research inventory, not a census of operating facilities or an authoritative site list.

## 1. Read all records with their source context

```{r retrieve_inventory}
library(tampaBayOpenData)

centers <- tbod_get_dataset(
  "tbrpc-regional-data-centers",
  fields = c("OBJECTID", "County", "Status", "Estimated_Size", "Data_Source"),
  total_timeout = 120
)
source <- tbod_provenance(centers)
stopifnot(isTRUE(source$complete), nrow(centers) == source$returned_rows)
print(data.frame(publisher = source$publisher, retrieved_at = source$retrieved_at,
                 records = nrow(centers), integrity = source$integrity))
```

## 2. Separate current facilities from other statuses

Counts are record counts in this particular research layer. The status labels are source values; this report does not independently verify an individual site's operating status.

```{r status_by_county, fig.width=7, fig.height=4}
clean_label <- function(x) {
  value <- trimws(as.character(x))
  value[is.na(value) | value == ""] <- "(missing)"
  value
}
county <- clean_label(centers$County)
status <- clean_label(centers$Status)
status_by_county <- table(county, status)
print(status_by_county)

barplot(colSums(status_by_county), col = "#267b83", border = NA,
        ylab = "Inventory records", xlab = "Source status",
        main = "Recorded status in this research inventory")
```

## 3. Audit the size field and cited sources

`Estimated_Size` is text and can contain different units. We classify its written unit rather than treating every number as comparable floor space. A URL in `Data_Source` indicates a citation is present; it does not validate the cited claim.

```{r audit_size_units}
size_text <- clean_label(centers$Estimated_Size)
size_unit <- ifelse(grepl("square feet", size_text, ignore.case = TRUE), "square feet",
             ifelse(grepl("acres?", size_text, ignore.case = TRUE), "acres", "other/missing"))
cited_url <- !is.na(centers$Data_Source) &
  grepl("^https?://", trimws(as.character(centers$Data_Source)))
print(table(status, size_unit))
print(data.frame(records = nrow(centers), records_with_source_url = sum(cited_url),
                 records_with_unparsed_size_unit = sum(size_unit == "other/missing")))
print(utils::head(data.frame(status, size = size_text, size_unit), 6))
```

## What this means

Status and measurement unit must be resolved before comparing capacities or summing sizes. The layer can guide follow-up research, but its small, selected inventory cannot support a countywide prevalence claim.

**Observed on 2026-10-06:** The 15-record inventory contained 13 records labeled Existing, one Rejected, and one Withdrawn. One written size was in acres and the other 14 in square feet. Rerunning this report will calculate the current figures.
