---
title: "Discovering Tampa Bay data"
description: "Search the checked catalog and live portals, then interpret and filter the results."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discovering Tampa Bay data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tampaBayOpenData` can search two related sources: a bundled catalog of 35
maintainer-checked layers, and public ArcGIS portal indexes operated by six
regional publishers. A checked layer has been configured and validated by the
package maintainers. A discovered layer is a candidate found in a live portal
index; its contents and reuse terms have not received the same validation.
The package does not host or mirror either source's records.

Live examples below are displayed but not run while the site is built. The
checked-catalog example runs offline.

## Start offline

```{r checked-catalog, eval=TRUE}
library(tampaBayOpenData)

tampaBayOpenData::tbod_list_portals()[, c("id", "publisher", "jurisdiction")]
tampaBayOpenData::tbod_list_datasets(source = "checked")[,
  c("id", "title", "jurisdiction", "validation_status")]
tampaBayOpenData::tbod_search_datasets(
  "park", source = "checked", spatial = TRUE,
  jurisdiction = c("stpete", "clearwater")
)
tampaBayOpenData::tbod_dataset_info("construction-permits")[
  c("title", "scope_note", "verified")]
```

`source = "checked"` reads the bundled registry without HTTP. Its IDs, such as
`construction-permits` and `stpete-parks`, are stable package names rather than
ArcGIS item IDs. `tbod_dataset_info(id)` reports each checked entry's scope, terms,
source URL, and endpoint-verification date. The latter is not a record-freshness
date. There are 20 Tampa checked layers and three each from St. Petersburg,
Clearwater, Hillsborough County, Pinellas County, and the Tampa Bay Regional
Planning Council. All six publishers also support live discovery.

## Discover live portal layers

```{r live-search}
found <- tbod_search_datasets("housing", source = "live", portals = "city")
found[, c("id", "title", "publisher", "validation_status", "modified")]

# A one-row result carries the source needed for inspection or retrieval.
selected <- found[1, ]
selected[, c("item_id", "service_url", "layer_id", "portal")]
tbod_dataset_info(selected)
```

By default, `tbod_list_datasets()` and `tbod_search_datasets()` use `source = "all"` and
search all six configured organizations, then overlay the checked registry.
`source = "live"` returns only layers found through portal indexes, including
checked layers when a live endpoint matches a checked entry. If a source is not
indexed or an index lags, it may not appear in live results; a compatible public
ArcGIS layer can still be retrieved by URL with `tbod_get_arcgis_layer()`.

Use `portals = "city"` for Tampa, `"tbrpc"` for the regional planning council,
or `"stpete"`, `"clearwater"`, `"hillsborough"`, or `"pinellas"` for the
other publishers. A vector selects several. `tbod_list_portals()` shows their
publisher names and source locations offline. `jurisdiction` defaults to
`"all"` for catalog functions and can be a code or a vector of codes. A
jurisdiction or publisher filter narrows the organizations scanned. Retrieval
and download functions default to `jurisdiction = NULL`: a checked ID resolves
across the bundled catalog when it is unique. If two jurisdictions use the
same ID, supply `jurisdiction` to select one.

`validation_status = "checked"` means maintainers validated the configured
layer. `"discovered"` means the item was found in a public portal index, not
that its contents or reuse terms were checked. Discovery keeps checked IDs,
scope, and terms when the live endpoint matches the registry, while using
current live tags, categories, and item modification time. Newly discovered
IDs have the form `arcgis:<item-id>:<layer-id>` and can be passed to
`tbod_get_dataset()` in a later online session. The item or its service can still
change or disappear. Retrieval validates query responses and requires `Query`
when a layer advertises a capabilities list.

## Search beyond the bundled publishers

`tbod_discover()` accepts a public ArcGIS sharing REST root, FeatureServer or
MapServer service root, or a layer URL. It uses the same discovery rows and
retrieval API without adding the source to the Tampa Bay catalog. For a portal,
`query` searches public service items; for a service URL, it filters layer
names. A queryable MapServer layer can be retrieved in the same way as a
FeatureServer layer. Portal search rows receive stable `arcgis:` IDs; rows
enumerated directly from a service use the layer URL as their ID.

```{r custom-discovery}
# Replace this example address with your portal's public sharing REST root.
portal <- "https://example.maps.arcgis.com/sharing/rest"
other <- tbod_discover(portal, query = "parks", max_items = 10)
if (nrow(other)) {
  data <- tbod_get_dataset(other[1, ], limit = 100)
  # A stable ArcGIS ID can also be resolved later through its own portal.
  same_layer <- tbod_get_dataset(other$id[[1]], portal = portal, limit = 100)
}
```

Custom sources default to an `"unspecified"` jurisdiction. Set `publisher`
and `jurisdiction` in `tbod_discover()` when you know them; these labels then
travel with the discovery row and its retrieval provenance. The `portal`
argument on retrieval applies only to a stable `arcgis:<item-id>:<layer-id>`
ID. When retrieving a direct layer URL, `tbod_get_dataset()` does not accept
a `jurisdiction` override.

## Filter results

Both catalog functions accept these optional filters:

| Filter | Match |
| --- | --- |
| `jurisdiction` | One or more codes from `tbod_list_portals()`; default `"all"`. |
| `publisher` | Exact publisher names from `tbod_list_portals()`, ignoring case. |
| `validation_status` | `"checked"` or `"discovered"`. |
| `spatial` | `TRUE` for geometry layers, `FALSE` for tables; unknown geometry is excluded when filtering. |
| `topic` | Literal substring in tags and categories, ignoring case. |
| `category` | Exact category path or final category label, ignoring case. |
| `modified_after`, `modified_before` | Inclusive bounds on the catalog's `modified` timestamp. |

Character filters accept vectors: values within one filter are alternatives;
separate filters and the search text must all match. Date bounds accept a
`Date` or strict `"YYYY-MM-DD"` string for a whole UTC day, or a `POSIXct`
value for an exact instant. Rows with unknown modification times are excluded
only when a date bound is supplied.

```{r county-search}
county_transport <- tbod_search_datasets(
  "road", source = "live", jurisdiction = "pinellas", spatial = TRUE,
  topic = "transportation", modified_after = as.Date("2025-01-01"),
  max_items = 10
)
county_transport[, c("id", "title", "publisher", "modified", "modified_source")]
```

`modified_source = "portal_item"` identifies a live ArcGIS item timestamp;
`"checked_snapshot"` identifies an upstream timestamp saved in the checked
registry. Neither establishes when every record was last updated. The
separate checked `verified` date records an endpoint check, not feature
freshness.

## Understand scan limits and failures

`max_items` caps **portal service items per organization**, not the number of
expanded layers. The default scans at most 25 items per organization. Use
`max_items = Inf` explicitly for a full live listing, with a suitable
`total_timeout`. A capped listing can miss matching items beyond the scan
limit. `jurisdiction` and `publisher` narrow organizations before scanning;
other filters apply after discovery and the checked overlay. Consequently,
an item matching `topic` or `modified_after` may be missed by a capped scan.

```{r scan-metadata}
results <- tbod_list_datasets(source = "live", portals = "stpete", max_items = 10)
attr(results, "discovery_complete")
attr(results, "discovery_truncated_portals")
attr(results, "discovery_scanned_items")
attr(results, "discovery_issues")
attr(results, "discovery_failed_portals")
```

Inspect `discovery_complete` before treating a result as a complete portal
catalog. Truncation and scanned-item attributes describe how far each scan
went. Skipped items and failed portals are reported in the issue and failure
attributes. If one portal fails, a combined search can still use other
publishers or rows collected before a later failure. If every selected live
portal fails before yielding rows, `source = "live"` errors; the default
combined search warns and returns checked rows. The checked
catalog remains available offline with `source = "checked"`.

Continue with [retrieval and downloads](retrieval.html) or read about
[coverage and source limits](coverage.html).
