---
title: "Discover and retrieve City of Tampa open data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discover and retrieve City of Tampa open data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tampaBayOpenData` helps you discover a curated catalog of City of Tampa open
datasets and query the City's public services through stable package IDs and a
consistent interface. It retrieves records as R tables or optional `sf`
objects, attaching provenance and completeness information to each retrieval.
The City publishes the data; this package helps you retrieve and assess it.

Catalog examples run offline. Examples that query City services are displayed
but not executed during vignette builds; run them in an internet-connected R
session.

## Find a dataset

```{r catalog, eval=TRUE}
library(tampaBayOpenData)
list_datasets()[, c("id", "title")]
search_datasets("permit")
```

Only `jurisdiction = "tampa"` is supported in version 0.1.0. Catalog IDs are
stable package identifiers, not ArcGIS item or layer IDs.

## Inspect dataset metadata

```{r metadata, eval=TRUE}
info <- dataset_info("construction-permits")
info[c("title", "source_url", "scope_note", "verified")]
```

`source_url` links to the City source, and `scope_note` describes limits on
what the dataset represents. The `verified` date records an endpoint check,
not the freshness of individual records. To inspect current field names,
request the live schema:

```{r live-metadata}
dataset_info("construction-permits", refresh = TRUE)$fields
```

## Retrieve records

```{r permits}
permits <- get_dataset(
  "construction-permits",
  where = "PROJECTSTATUS IS NOT NULL",
  fields = c("RECORD_ID", "PROJECTSTATUS", "LASTUPDATE"),
  limit = 100
)
head(permits)
```

`where` uses ArcGIS SQL, and `fields` uses exact source names. By default,
`limit = Inf` retrieves every match. A finite limit returns a subset and
provenance marks it incomplete if more records match. `order_by` requests
server ordering where supported; `get_permits()` is a shortcut for this
dataset. See `?get_dataset` for other filters and limits.

Results retain source field names and strings. Declared date fields become
UTC `POSIXct` when their time-zone meaning is known. Dates with unknown
time-zone semantics remain raw milliseconds with a warning.

## Explore the results

```{r permit-statuses}
sort(table(permits$PROJECTSTATUS), decreasing = TRUE)
```

These counts describe the retrieved subset, not every matching permit.

## Work with spatial data

The same interface returns `sf` objects with `spatial = TRUE`. Install the
optional `sf` package if needed. After loading `tampaBayOpenData` in the first
example, run the next three chunks in order in an internet-connected R session.

```{r spatial}
# Request one CRS for both layers so their coordinates can share a map.
projects <- get_dataset(
  "development-cases",
  spatial = TRUE,
  out_sr = 4326
)

# Only the boundary ID is needed; spatial = TRUE still returns its geometry.
boundary <- get_dataset(
  "city-boundary",
  fields = "OBJECTID",
  spatial = TRUE,
  out_sr = 4326
)
# Check the CRS reported for the case locations.
sf::st_crs(projects)$epsg
```

Both results are `sf` objects in WGS 84 (EPSG:4326). The CRS line confirms the
value for `projects`. Omitting `out_sr` keeps each layer's native CRS; point,
line, and polygon layers use the same API.

Records without geometry remain in `projects`. Make a subset of locations to
plot:

```{r spatial-prepare}
# Preserve projects; create a plotting subset with nonempty geometries.
mapped_projects <- projects[!sf::st_is_empty(projects), ]
```

Draw the city boundary for context, then add the case locations:

```{r spatial-plot, fig.width=7, fig.height=5.5}
# Draw the polygon first so it forms a backdrop for the case locations.
plot(
  sf::st_geometry(boundary),
  col = "#EDF4F6",
  border = "#7898A5",
  lwd = 1.2,
  axes = TRUE,
  graticule = TRUE,
  col_graticule = "#DCE7EB",
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Active development cases in Tampa"
)

# Filled circles have separate outlines and translucent fills.
if (nrow(mapped_projects) > 0L) {
  case_fill <- grDevices::adjustcolor("#E76F51", alpha.f = 0.7)
  plot(
    sf::st_geometry(mapped_projects),
    add = TRUE,
    pch = 21,
    bg = case_fill,
    col = "#9B4934",
    cex = 0.85
  )

  # Match the legend symbol to the plotted points.
  graphics::legend(
    "bottomleft",
    legend = "Development case location",
    pch = 21,
    pt.bg = case_fill,
    col = "#9B4934",
    bg = "white",
    box.col = "#DCE7EB",
    cex = 0.85
  )
}
```

The `city-boundary` layer is adjusted to follow the shoreline. It provides map
context, not a legal survey boundary.

## Keep track of where your data came from

```{r provenance}
source <- dataset_provenance(projects)
saveRDS(
  list(data = projects, provenance = source),
  "development-cases.rds"
)
```

Provenance includes the publisher, source URL, query endpoint, retrieval time,
row counts, and completeness. Save it separately with your data because later
R transformations may discard custom attributes.

## Important limitations

The permit and development-case layers are active views, not necessarily
complete historical ledgers. Read `dataset_info(id)$scope_note` and
`dataset_info(id)$terms` before interpreting their coverage or reuse terms.

Live ArcGIS data can change during a multi-request retrieval. The client
checks fetched IDs against the matching count and object-ID manifest where
applicable; these checks cannot freeze every attribute at one instant.

If a field is uncertain, inspect `dataset_info(id, refresh = TRUE)$fields`.
For a temporary service failure, check the source URL and retry. Queries
matching more than one million records are rejected because a complete
object-ID manifest is required; narrow `where` first.

When using City data for analysis, keep its provenance and revisit the
dataset's stated scope before sharing conclusions.
