---
title: "Discover and retrieve Tampa Bay ArcGIS data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discover and retrieve Tampa Bay ArcGIS data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tampaBayOpenData` searches public City of Tampa and Tampa Bay Regional Planning
Council ArcGIS datasets. It also includes 20 City layers that package maintainers
have checked. The same client retrieves checked or newly discovered layers as R
tables or optional `sf` objects, with provenance and completeness information.

The checked catalog examples run offline. Live discovery and retrieval examples
are displayed but not executed during vignette builds; run them in an R session
with internet access.

## Find a dataset

```{r catalog, eval=TRUE}
library(tampaBayOpenData)
list_datasets(source = "checked")[, c("id", "title", "validation_status")]
search_datasets("permit", source = "checked")
```

`source = "checked"` reads the bundled City registry without a network request.
Checked IDs are stable package names, separate from ArcGIS item and layer IDs.
The `jurisdiction` argument selects checked City entries; `portals` selects
live City and regional organizations.

Search the live portals when you need more than the checked catalog:

```{r live-discovery}
# The default `source = "all", portals = "all"` also includes checked entries.
found <- search_datasets("housing", source = "live", portals = "city")
found[, c("id", "title", "publisher", "geometry_type", "validation_status")]

# Inspect one candidate's source before retrieving it.
selected <- found[1, ]
selected[, c("item_id", "service_url", "layer_id", "portal")]
```

`validation_status = "checked"` means maintainers specifically tested that
layer. `"discovered"` identifies a candidate from a public portal index;
maintainers have not validated its contents or reuse terms. Retrieval checks
that the selected layer advertises ArcGIS Query capability. Unchecked live IDs
have the form `arcgis:<item-id>:<layer-id>` and remain usable if a title changes.
A portal item may still be deleted or its service
changed. `list_datasets()` searches both public organizations by default;
`source = "checked"` remains available if the network is unavailable.

```{r discovered-retrieval}
# A one-row discovery result carries its source and item metadata.
housing <- get_dataset(selected, limit = 100)
dataset_provenance(housing)[c("item_id", "portal", "validation_status")]

# Its ID can also be passed to get_dataset() in a later online session:
# housing <- get_dataset(selected$id[[1]], limit = 100)
```

The discovery result includes the item's `modified` date, tags, and service
location when available. That date is not necessarily the last update to its
records. `max_items` limits portal items scanned per organization, not layers.

## Inspect dataset metadata

```{r metadata, eval=TRUE}
info <- dataset_info("construction-permits")
info[c("title", "source_url", "scope_note", "verified")]
```

`source_url` links to the checked City source, and `scope_note` describes limits
on what the dataset represents. The `verified` date records an endpoint check,
not the freshness of individual records. To inspect current field names,
request the live schema:

```{r live-metadata}
dataset_info("construction-permits", refresh = TRUE)$fields
```

## Retrieve records

```{r permits}
permits <- get_dataset(
  "construction-permits",
  where = "PROJECTSTATUS IS NOT NULL",
  fields = c("RECORD_ID", "PROJECTSTATUS", "LASTUPDATE"),
  limit = 100
)
head(permits)
```

`where` uses ArcGIS SQL rather than Socrata SoQL, and `fields` uses exact source
names. By default, `limit = Inf` retrieves every match. A finite limit returns a
subset, and provenance marks it incomplete if more records match. `order_by`
requests server ordering where supported; `get_permits()` is a shortcut for this
dataset. See `?get_dataset` for other filters and limits.

Results retain source field names and strings. Declared date fields become
UTC `POSIXct` when their time-zone meaning is known. Dates with unknown
time-zone semantics remain raw milliseconds with a warning.

You can also retrieve a compatible public ArcGIS layer directly by its URL:

```{r direct-layer}
direct <- get_arcgis_layer(
  "https://arcgis.tampagov.net/arcgis/rest/services/OpenData/SolidWaste/MapServer/4",
  limit = 10
)
dataset_provenance(direct)$validation_status # "not_checked" for direct URL access
```

## Explore the results

```{r permit-statuses}
sort(table(permits$PROJECTSTATUS), decreasing = TRUE)
```

These counts describe the retrieved subset, not every matching permit.

## Work with spatial data

The same interface returns `sf` objects with `spatial = TRUE`. Install the
optional `sf` package if needed. After loading `tampaBayOpenData` in the first
example, run the next three chunks in order in an internet-connected R session.

```{r spatial}
# Request one CRS for both layers so their coordinates can share a map.
projects <- get_dataset(
  "development-cases",
  spatial = TRUE,
  out_sr = 4326
)

# Only the boundary ID is needed; spatial = TRUE still returns its geometry.
boundary <- get_dataset(
  "city-boundary",
  fields = "OBJECTID",
  spatial = TRUE,
  out_sr = 4326
)
# Check the CRS reported for the case locations.
sf::st_crs(projects)$epsg
```

Both results are `sf` objects in WGS 84 (EPSG:4326). The CRS line confirms the
value for `projects`. Omitting `out_sr` keeps each layer's native CRS; point,
line, and polygon layers use the same API.

Records without geometry remain in `projects`. Make a subset of locations to
plot:

```{r spatial-prepare}
# Preserve projects; create a plotting subset with nonempty geometries.
mapped_projects <- projects[!sf::st_is_empty(projects), ]
```

Draw the city boundary for context, then add the case locations:

```{r spatial-plot, fig.width=7, fig.height=5.5}
# Draw the polygon first so it forms a backdrop for the case locations.
plot(
  sf::st_geometry(boundary),
  col = "#EDF4F6",
  border = "#7898A5",
  lwd = 1.2,
  axes = TRUE,
  graticule = TRUE,
  col_graticule = "#DCE7EB",
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Active development cases in Tampa"
)

# Filled circles have separate outlines and translucent fills.
if (nrow(mapped_projects) > 0L) {
  case_fill <- grDevices::adjustcolor("#E76F51", alpha.f = 0.7)
  plot(
    sf::st_geometry(mapped_projects),
    add = TRUE,
    pch = 21,
    bg = case_fill,
    col = "#9B4934",
    cex = 0.85
  )

  # Match the legend symbol to the plotted points.
  graphics::legend(
    "bottomleft",
    legend = "Development case location",
    pch = 21,
    pt.bg = case_fill,
    col = "#9B4934",
    bg = "white",
    box.col = "#DCE7EB",
    cex = 0.85
  )
}
```

The `city-boundary` layer is adjusted to follow the shoreline. It provides map
context, not a legal survey boundary.

## Keep track of where your data came from

```{r provenance}
source <- dataset_provenance(projects)
saveRDS(
  list(data = projects, provenance = source),
  "development-cases.rds"
)
```

Provenance includes the publisher, source URL, portal and item ID when known,
validation status, query endpoint, retrieval time, row counts, and
completeness. Save it separately with your data because later R transformations
may discard custom attributes.

## Important limitations

The permit and development-case layers are active views, not necessarily
complete historical ledgers. Read `dataset_info(id)$scope_note` and
`dataset_info(id)$terms` before interpreting their coverage or reuse terms.

Live ArcGIS data can change during a multi-request retrieval. The client
checks fetched IDs against the matching count and object-ID manifest where
applicable; these checks cannot freeze every attribute at one instant.

If a field is uncertain, inspect `dataset_info(id, refresh = TRUE)$fields`.
For a temporary service failure, check the source URL and retry. Queries
matching more than one million records are rejected because a complete
object-ID manifest is required; narrow `where` first.

Discovered items may lack clear reuse terms or may be stale in the portal
index. Review each item's source and terms before analysis; the package does
not verify those terms. The package's MIT
license applies to its code, not to data from either source organization.
