---
title: "Discover and retrieve City of Tampa open data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discover and retrieve City of Tampa open data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tampaBayOpenData` connects a small catalog of verified City of Tampa sources to
ordinary R tables and spatial objects. The City publishes the data; this package
provides discovery, querying, pagination, parsing, and provenance.

Catalog examples below run offline. Retrieval examples are displayed without
contacting government services when this vignette builds. Run them in your R
session with an internet connection; install `sf` for the spatial example.

## Find a dataset

```{r load, eval=TRUE}
library(tampaBayOpenData)
```

The catalog is bundled with the package. Its IDs are stable package identifiers,
separate from upstream titles and ArcGIS layer numbers.

```{r catalog, eval=TRUE}
datasets <- list_datasets()
datasets
search_datasets("permit")
```

Only `jurisdiction = "tampa"` is implemented in v0.1. Explicit jurisdiction works
with the same API:

```{r explicit-jurisdiction, eval=TRUE}
search_datasets("planning", jurisdiction = "tampa")
```

## Inspect the source

```{r metadata, eval=TRUE}
info <- dataset_info("construction-permits")
str(info, max.level = 1)
```

Metadata identifies the City publisher, official landing page, service/layer,
geometry, useful date fields, terms, and endpoint verification date. Fetch live
metadata to inspect the current field schema and service capabilities:

```{r live-metadata}
dataset_info("construction-permits", refresh = TRUE)
```

The permit source contains permit locations and source attributes. A layer's
title does not by itself establish a complete permit or inspection history.

## Retrieve and analyze attributes

```{r retrieve-permits}
permits <- get_dataset("construction-permits")
head(permits)
table(permits$PROJECTSTATUS, useNA = "ifany")
```

The default returns a tibble and retrieves every matching record with automatic
pagination. Source names such as `RECORD_ID` and `PROJECTSTATUS` retain their
spelling and case. Declared date fields such as `LASTUPDATE` become UTC `POSIXct`
where their semantics are known. Date-looking strings remain strings.

Reduce the downloaded columns and filter at the service:

```{r filter-permits}
selected <- get_dataset(
  "construction-permits",
  where = "PROJECTSTATUS IS NOT NULL",
  fields = c("RECORD_ID", "PROJECTSTATUS", "LASTUPDATE")
)
head(selected)
```

`where` is an ArcGIS SQL expression. Use source field names and consult the live
metadata when a service rejects a field. Advanced ArcGIS spatial/time/version
parameters can be supplied through `query`; see `?get_dataset` for supported
parameters.

For a deliberate subset ordered by reported update time:

```{r recent-permits}
recent <- get_permits(order_by = "LASTUPDATE DESC", limit = 100)
head(recent)
dataset_provenance(recent)
```

The convenience function calls the same generic client. A limit below the
matching count is recorded with a false completeness flag. It is suitable for
exploration, but is not the full matching dataset.

## Retrieve spatial development cases

The verified `development-cases` dataset represents active entitlement locations.
Its source fields include `RECORDID`, `ADDRESS`, `APPSTATUS`, and `LASTUPDATE`.

```{r spatial-development}
projects <- get_dataset(
  "development-cases",
  fields = c("RECORDID", "ADDRESS", "APPSTATUS", "LASTUPDATE"),
  spatial = TRUE,
  out_sr = 4326
)
sf::st_crs(projects)
head(projects)
plot(sf::st_geometry(projects))
```

`out_sr = 4326` requests WGS 84 coordinates from the service. Omit it to retain
the source spatial reference. The client reads CRS metadata rather than assigning
one shared projection to all datasets. Missing geometry remains represented in
the returned `sf` object.

Polygon data use the same API:

```{r boundary}
boundary <- get_dataset("city-boundary", spatial = TRUE, out_sr = 4326)
plot(sf::st_geometry(boundary))
```

## Keep provenance with an analysis

```{r provenance}
provenance <- dataset_provenance(projects)
provenance
saveRDS(list(data = projects, provenance = provenance), "development-cases.rds")
```

Provenance records the government source, package dataset ID, endpoint, retrieval
time, query options, expected/returned rows, and completeness. Save it separately
if subsequent transformations drop custom attributes. Do not cite this package
as the government publisher.

The client compares a matching count and object-ID manifest with downloaded
batches. A truncated response, unexpected ID, missing record, or duplicate page
produces an error rather than a result presented as complete. Live services can
change during a request; those checks establish record membership, not a frozen
transactional snapshot of every attribute.

## Handle service changes

An unknown ID is resolved with `list_datasets()`. An unknown field is resolved
with `dataset_info(id, refresh = TRUE)`. For an unavailable service, inspect the
dataset's official source page, retry later, or increase `timeout`:

```{r longer-timeout}
capital <- get_capital_projects(timeout = 60)
```

For a retrieval-integrity failure, retry or narrow `where`; do not silently
discard the failure. Very large queries can exceed the upstream object-ID
manifest cap and must be filtered. Spatial retrieval requires `sf` and a
resolvable source or requested CRS. Unknown date time zones remain raw with a
warning.

Read the source's limitations and terms before using a dataset. The catalog
verification date describes an endpoint check, not the freshness or completeness
of every government record.
