---
title: "Discover and retrieve Tampa Bay ArcGIS data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discover and retrieve Tampa Bay ArcGIS data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tampaBayOpenData` searches public ArcGIS datasets from Tampa, St. Petersburg,
Clearwater, Hillsborough County, Pinellas County, and the Tampa Bay Regional
Planning Council. It also includes 26 layers across the three cities that
package maintainers have checked.
The same client retrieves checked or newly discovered layers as R
tables or optional `sf` objects, with provenance and completeness information.

The checked catalog examples run offline. Live discovery and retrieval examples
are displayed but not executed during vignette builds; run them in an R session
with internet access.

## Find a dataset

```{r catalog, eval=TRUE}
library(tampaBayOpenData)
list_portals()
list_datasets(source = "checked")[,
  c("id", "title", "jurisdiction", "validation_status")]
search_datasets("park", source = "checked", spatial = TRUE,
                jurisdiction = c("stpete", "clearwater"))
```

`source = "checked"` reads the bundled registry without a network request.
Checked IDs are stable package names, separate from ArcGIS item and layer IDs.
For catalog functions, `jurisdiction = "all"` is the default. A code or vector
filters both checked and live results: `"tampa"`, `"tampa-bay"`, `"stpete"`,
`"clearwater"`, `"hillsborough"`, or `"pinellas"`. The `portals` argument selects live
organizations: `"city"` (Tampa), `"tbrpc"`, `"stpete"`, `"clearwater"`,
`"hillsborough"`, `"pinellas"`, or `"all"` (default). `list_portals()` lists the
configured publishers and their portal locations offline. The county portals
extend live discovery; the checked catalog still contains 26 city layers.
Selecting a county with `source = "checked"` returns an empty catalog.
Retrieval and download functions retain their default `jurisdiction = "tampa"`;
omitting it also resolves uniquely named checked IDs across cities.

Search the live portals when you need more than the checked catalog:

```{r live-discovery}
# The default `source = "all", portals = "all"` also includes checked entries.
found <- search_datasets("housing", source = "live", portals = "city")
found[, c("id", "title", "publisher", "geometry_type", "validation_status")]

# Inspect one candidate's source before retrieving it.
selected <- found[1, ]
selected[, c("item_id", "service_url", "layer_id", "portal")]
```

`validation_status = "checked"` means maintainers specifically tested that
layer. `"discovered"` identifies a candidate from a public portal index;
maintainers have not validated its contents or reuse terms. Retrieval validates
query responses and requires `Query` when the layer declares its capabilities.
Unchecked live IDs
have the form `arcgis:<item-id>:<layer-id>` and remain usable if a title changes.
A portal item may still be deleted or its service
changed. `list_datasets()` searches all six public organizations by default;
`source = "checked"` remains available if the network is unavailable.

```{r discovered-retrieval}
# A one-row discovery result carries its source and item metadata.
housing <- get_dataset(selected, limit = 100)
dataset_provenance(housing)[c("item_id", "portal", "validation_status")]

# Its ID can also be passed to get_dataset() in a later online session:
# housing <- get_dataset(selected$id[[1]], limit = 100)
```

The discovery result includes the item's `modified` date, tags, and service
location when available. That date is not necessarily the last update to its
records. `max_items` limits portal items scanned per organization, not layers;
it defaults to 25. Set `max_items = Inf` for a full scan. Inspect the catalog's
`discovery_complete`, `discovery_truncated_portals`, and `discovery_scanned_items`
attributes before treating a bounded listing as a complete catalog.

### Narrow a search

Both `list_datasets()` and `search_datasets()` accept optional filters:

```{r county-search}
county_transport <- search_datasets(
  "road", source = "live", jurisdiction = "pinellas", spatial = TRUE,
  topic = "transportation", modified_after = as.Date("2025-01-01"),
  max_items = 10
)
county_transport[, c("id", "title", "publisher", "modified", "modified_source")]
```

`publisher` matches exact names from `list_portals()`, ignoring case;
`validation_status` selects `"checked"` or `"discovered"`. Live-only results can
still contain checked matches. `spatial = TRUE` selects geometry layers,
`FALSE` selects tables, and either filter excludes unknown geometry.
`topic` matches literal substrings in tags and categories, ignoring case.
`category` matches a complete category path or its final label, ignoring case.
Character filters accept vectors; values in one filter are alternatives, while
different filters must all match.

`modified_after` and `modified_before` are inclusive bounds on `modified`.
Use `Date` values or strict `"YYYY-MM-DD"` strings to include whole UTC days,
or `POSIXct` values for exact instants. Rows with unknown item dates are dropped
when a date bound is supplied. `modified_source = "portal_item"` identifies a
live item timestamp; `"checked_snapshot"` identifies an upstream timestamp
saved in the checked registry. Neither establishes record freshness, and the
checked `verified` date remains a separate endpoint check.

Jurisdiction and publisher filters reduce the organizations scanned. Other
filters run after discovery and the checked overlay, so a bounded `max_items`
scan can miss matching items beyond that limit. Inspect discovery attributes
and increase the item limit when a broader search is needed. When a live layer
matches a checked endpoint, the result keeps its checked identity and terms
alongside current live tags, categories, and item modification date.

## Inspect dataset metadata

```{r metadata, eval=TRUE}
info <- dataset_info("construction-permits")
info[c("title", "source_url", "scope_note", "verified")]
```

`source_url` links to the checked City source, and `scope_note` describes limits
on what the dataset represents. The `verified` date records an endpoint check,
not the freshness of individual records. To inspect current field names,
request the live schema:

```{r live-metadata}
dataset_info("construction-permits", refresh = TRUE)$fields
```

## Retrieve records

```{r permits}
permits <- get_dataset(
  "construction-permits",
  where = "PROJECTSTATUS IS NOT NULL",
  fields = c("RECORD_ID", "PROJECTSTATUS", "LASTUPDATE"),
  limit = 100
)
head(permits)
```

`where` uses ArcGIS SQL rather than Socrata SoQL, and `fields` uses exact source
names. By default, `limit = Inf` retrieves every match. A finite limit returns a
subset, and provenance marks it incomplete if more records match. `order_by`
requests server ordering where supported; `get_permits()` is a shortcut for this
dataset. See `?get_dataset` for other filters and limits.

For sources with more than 10,000 matches and reliable ordering and pagination,
finite previews of at most 1,000 rows verify the returned IDs with a subset
manifest. Other retrievals use a full manifest; set `integrity = "full"` to
request that explicitly. Provenance identifies `subset-manifest`,
`full-manifest`, or `count-only` for a zero-row limit. A finite preview is marked
incomplete when more records match.

`timeout` limits each HTTP attempt. The overall `total_timeout` defaults to 120
seconds for retrieval, 90 seconds for catalog discovery, and 60 seconds for
metadata inspection. Set `total_timeout = Inf` explicitly for a long operation.
Requests and retry waits use the remaining budget; parsing is checked after
it finishes and can exceed the budget before reporting a timeout.

Results retain source field names and text fields. ArcGIS epoch-millisecond
date fields become UTC `POSIXct` unless the source declares an unknown time zone.
In that case they remain raw milliseconds with a warning; UTC editor-tracking
creation and edit dates identified by the layer metadata are still converted.
Date-only fields become `Date`; time-only and timestamp-offset fields remain
strings.

You can also retrieve a compatible public ArcGIS layer directly by its URL:

```{r direct-layer}
direct <- get_arcgis_layer(
  "https://arcgis.tampagov.net/arcgis/rest/services/OpenData/SolidWaste/MapServer/4",
  limit = 10
)
dataset_provenance(direct)$validation_status # "not_checked" for direct URL access
```

For larger retrievals, write resumable chunks to a persistent directory outside
the package source and read one chunk at a time:

```{r download}
saved <- download_dataset(
  "construction-permits",
  path = "~/tbOpenData-downloads/permits",
  fields = c("OBJECTID", "RECORD_ID", "LASTUPDATE"),
  page_size = 500,
  total_timeout = 600
)
saved$complete
first_chunk <- readRDS(saved$files[[1]])
dataset_provenance(first_chunk)
```

Repeat the same download call to resume. The client rechecks current metadata and
the full matching ID manifest, validates saved chunk checksums and provenance,
and skips completed chunks. Changed query options, schema, CRS, or ID membership
require a new directory. Chunk attributes can vary across requests or resumes,
so the files do not form a frozen snapshot. Each chunk has its own completeness
record; `saved$complete` describes the whole matching download. This path keeps
only bounded feature batches in memory and retains the one-million-ID manifest
limit. Download operations default to `total_timeout = Inf`; a finite deadline
leaves completed chunks available for the next call.

An exclusive directory lock prevents concurrent writers. A terminated R process
can leave an empty `.lock` directory: confirm that process has stopped before
removing the lock and resuming. `download_arcgis_layer()` supports the same
chunked path for direct URLs.

## Explore the results

```{r permit-statuses}
sort(table(permits$PROJECTSTATUS), decreasing = TRUE)
```

These counts describe the retrieved subset, not every matching permit.

## Work with spatial data

The same interface returns `sf` objects with `spatial = TRUE`. Install the
optional `sf` package if needed. After loading `tampaBayOpenData` in the first
example, run the next three chunks in order in an internet-connected R session.

```{r spatial}
# Request one CRS for both layers so their coordinates can share a map.
projects <- get_dataset(
  "development-cases",
  spatial = TRUE,
  out_sr = 4326
)

# Only the boundary ID is needed; spatial = TRUE still returns its geometry.
boundary <- get_dataset(
  "city-boundary",
  fields = "OBJECTID",
  spatial = TRUE,
  out_sr = 4326
)
# Check the CRS reported for the case locations.
sf::st_crs(projects)$epsg
```

Both results are `sf` objects in WGS 84 (EPSG:4326). The CRS line confirms the
value for `projects`. Omitting `out_sr` keeps each layer's native CRS; point,
line, and polygon layers use the same API.

Records without geometry remain in `projects`. Make a subset of locations to
plot:

```{r spatial-prepare}
# Preserve projects; create a plotting subset with nonempty geometries.
mapped_projects <- projects[!sf::st_is_empty(projects), ]
```

Draw the city boundary for context, then add the case locations:

```{r spatial-plot, fig.width=7, fig.height=5.5}
# Draw the polygon first so it forms a backdrop for the case locations.
plot(
  sf::st_geometry(boundary),
  col = "#EDF4F6",
  border = "#7898A5",
  lwd = 1.2,
  axes = TRUE,
  graticule = TRUE,
  col_graticule = "#DCE7EB",
  xlab = "Longitude",
  ylab = "Latitude",
  main = "Active development cases in Tampa"
)

# Filled circles have separate outlines and translucent fills.
if (nrow(mapped_projects) > 0L) {
  case_fill <- grDevices::adjustcolor("#E76F51", alpha.f = 0.7)
  plot(
    sf::st_geometry(mapped_projects),
    add = TRUE,
    pch = 21,
    bg = case_fill,
    col = "#9B4934",
    cex = 0.85
  )

  # Match the legend symbol to the plotted points.
  graphics::legend(
    "bottomleft",
    legend = "Development case location",
    pch = 21,
    pt.bg = case_fill,
    col = "#9B4934",
    bg = "white",
    box.col = "#DCE7EB",
    cex = 0.85
  )
}
```

The `city-boundary` layer is adjusted to follow the shoreline. It provides map
context, not a legal survey boundary.

## Keep track of where your data came from

```{r provenance}
source <- dataset_provenance(projects)
saveRDS(
  list(data = projects, provenance = source),
  "development-cases.rds"
)
```

Provenance includes the publisher, source URL, portal and item ID when known,
validation status, query endpoint, retrieval time, row counts, and
completeness. Save it separately with your data because later R transformations
may discard custom attributes.

## Important limitations

The permit and development-case layers are active views, not necessarily
complete historical ledgers. Read `dataset_info(id)$scope_note` and
`dataset_info(id)$terms` before interpreting their coverage or reuse terms.

Live ArcGIS data can change during a multi-request retrieval. The client
checks fetched IDs against the matching count and object-ID manifest where
applicable; these checks cannot freeze every attribute at one instant.

If a field is uncertain, inspect `dataset_info(id, refresh = TRUE)$fields`.
For a temporary service failure, check the source URL and retry. Full-manifest
queries matching more than one million records are rejected; narrow `where`
first. Eligible finite previews can verify a bounded subset instead.

Discovered items may lack clear reuse terms or may be stale in the portal
index. Review each item's source and terms before analysis; the package does
not verify those terms. The package's MIT
license applies to its code, not to data from either source organization.
