---
title: "Get started with Tampa Bay open data"
description: "Find a dataset, retrieve its records, work with spatial data, and trace its source."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with Tampa Bay open data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tampaBayOpenData` gives R users one interface for finding and retrieving
public ArcGIS data from Tampa, St. Petersburg, Clearwater, Hillsborough
County, Pinellas County, and the Tampa Bay Regional Planning Council. It
bundles 35 maintainer-checked layers and also discovers public layers
from the six live organizations. Results are R tibbles or optional `sf`
objects with source provenance and completeness information.

The offline catalog examples below run during the vignette build. Live
discovery and retrieval examples are displayed but not run; try them in an
internet-connected R session.

## Find data

Start with the checked catalog when you want an offline, reproducible list of
package IDs:

```{r checked-catalog, eval=TRUE}
library(tampaBayOpenData)

tampaBayOpenData::list_portals()[, c("id", "publisher", "jurisdiction")]
tampaBayOpenData::search_datasets("park", source = "checked")[,
  c("id", "title", "jurisdiction", "validation_status")]
tampaBayOpenData::dataset_info("construction-permits")[
  c("title", "scope_note", "verified")]
```

`source = "checked"` uses a bundled registry without making a network
request. Checked IDs, including `stpete-parks` and `construction-permits`,
remain stable package names. `dataset_info()` reports the source, terms, and
scope of a checked entry. For more choices, search live portal indexes:

```{r live-discovery}
found <- search_datasets("housing", source = "live", portals = "city")
found[, c("id", "title", "publisher", "validation_status")]
```

Catalog searches scan all six publishers by default. Set `portals` or
`jurisdiction` to narrow them. `checked` means the package maintainers
validated the configured layer; `discovered` means it appeared in a live
portal index but has not received the same validation. Inspect a discovered
source and its reuse terms before analysis. See [discovering Tampa Bay
data](discovery.html) for all filters, live scan limits, and discovery status.

## Retrieve records

Pass a checked ID, a discovered stable ID, or a one-row search result to
`get_dataset()`. This example requests a small subset of a checked Tampa
layer as a tibble:

```{r retrieve}
permits <- get_dataset(
  "construction-permits",
  where = "PROJECTSTATUS IS NOT NULL",
  fields = c("RECORD_ID", "PROJECTSTATUS", "LASTUPDATE"),
  limit = 100
)
head(permits)
sort(table(permits$PROJECTSTATUS), decreasing = TRUE)
```

The counts describe the retrieved subset, not all matching permits. Omit
`limit` to retrieve all matching rows. A discovered result can travel to
retrieval with its source information:

```{r discovered-retrieval}
housing <- get_dataset(found[1, ], limit = 100)
dataset_provenance(housing)[c("publisher", "validation_status", "complete")]
```

## Work with spatial data

Install the optional `sf` package and request `spatial = TRUE` for an `sf`
result. `out_sr = 4326` asks the service for WGS 84 coordinates; without
`out_sr`, the client keeps the native CRS.

```{r spatial}
projects <- get_dataset("development-cases", spatial = TRUE,
                        out_sr = 4326)
sf::st_crs(projects)$epsg
```

The [retrieval article](retrieval.html) shows a map with this layer and the
Tampa boundary. It also explains ArcGIS filters, pagination, integrity
checks, time budgets, data types, direct layer URLs, and resumable downloads.

## Trace the source

Provenance records where the result came from, the query, retrieval time,
row counts, validation status, integrity method, and whether the result is
complete. Save it separately before transforming the data, because some R
operations may discard attached attributes.

```{r provenance}
source <- dataset_provenance(projects)
saveRDS(list(data = projects, provenance = source),
        "development-cases.rds")
```

Read [coverage, provenance, and source limits](coverage.html) before treating
a layer as a complete historical record or interpreting a map as a legal
boundary. That article also explains source attribution, citation, government
data reuse terms, and monitoring evidence. The package is independent of
the data publishers and rOpenSci.
