---
title: "Retrieving and downloading data"
description: "Retrieve and download records or spatial layers with integrity checks."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieving and downloading data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Use `tbod_get_dataset()` for a checked package ID, a one-row discovery result, or a
stable discovered ID. Use `tbod_get_arcgis_layer()` when you already have a
compatible public HTTPS ArcGIS layer URL. Both paths use the same retrieval,
type conversion, integrity checks, and provenance. These examples contact live
publishers when run in an R session; none run during the site build.

## Retrieve and filter records

```{r permits}
library(tampaBayOpenData)

permits <- tbod_get_dataset(
  "construction-permits",
  where = "PROJECTSTATUS IS NOT NULL",
  fields = c("RECORD_ID", "PROJECTSTATUS", "LASTUPDATE"),
  limit = 100
)
head(permits)
tbod_provenance(permits)[c("matched_rows", "returned_rows", "complete", "integrity")]
```

`where` uses ArcGIS SQL, not Socrata SoQL, and `fields` requires exact source
field names. Inspect current fields with
`tbod_dataset_info("construction-permits", refresh = TRUE)$fields`; its metadata
also reports query capabilities. `query` accepts supported advanced ArcGIS
spatial, time, and version filters. It does not allow overriding parameters
that control the response format, field selection, aggregation, geometry
simplification, or pagination; see `?tbod_get_dataset` for the accepted names.

The default `limit = Inf` retrieves all matches. A finite limit deliberately
returns a subset and `tbod_provenance()` records `complete = FALSE` when
additional records match. Use `order_by` for a reproducible ordered subset
only when the service supports ordering and offset pagination; an object-ID
tie breaker is added. `limit = 0` returns a typed empty result after obtaining
the matching count. `tbod_get_permits()`, `tbod_get_development_cases()`, and
`tbod_get_capital_projects()` are short calls to `tbod_get_dataset()` for three checked
Tampa layers.

For a one-row discovery result, pass the row itself so its source and publisher
travel with the request:

```{r discovered-layer}
found <- tbod_search_datasets("housing", source = "live", portals = "city")
housing <- tbod_get_dataset(found[1, ], limit = 100)
tbod_provenance(housing)[c("publisher", "item_id", "validation_status")]
```

For a compatible direct layer, pass a URL ending in `/FeatureServer/<id>` or
`/MapServer/<id>`:

```{r direct-layer}
yard_waste <- tbod_get_arcgis_layer(
  "https://arcgis.tampagov.net/arcgis/rest/services/OpenData/SolidWaste/MapServer/4",
  limit = 10
)
tbod_provenance(yard_waste)$validation_status # "not_checked"
```

Direct URL access does not infer a publisher or ArcGIS portal item and has
`not_checked` validation status. It cannot take a `jurisdiction` override.
Inspect the service's source and reuse terms yourself.

## Check the live schema

Checked layers have a bundled field, geometry, CRS, object-ID, and endpoint
snapshot. `tbod_schema(id)` reads that snapshot offline;
`tbod_schema(id, refresh = TRUE)` reads the current layer metadata.
`tbod_check_schema(id)` compares them and returns a report:

```{r schema-check}
expected <- tbod_schema("construction-permits")
report <- tbod_check_schema("construction-permits")
report$status
report$issues[, c("category", "severity", "field", "expected", "actual")]
```

The status is `"unchanged"`, `"compatible"`, or `"incompatible"` for a
checked layer. Added fields and changed aliases are compatible drift: checked
retrieval warns and proceeds. Removed or renamed fields, changed field types,
object-ID fields, geometry, CRS, or endpoint are incompatible: checked
retrieval stops before querying records. A vanished endpoint is reported as
`endpoint_disappeared`. Inspect `report$issues` to identify the change.

A direct URL or discovered layer has no bundled expected schema. Save its
current metadata and supply it on a later check; an untracked check without
`expected` returns `status = "untracked"`. For a stable ID from a custom
Enterprise portal, also pass `portal` to `tbod_schema()` or
`tbod_check_schema()`.

```{r external-schema}
# Replace this example address with a public ArcGIS layer URL.
url <- "https://example.org/arcgis/rest/services/Example/FeatureServer/0"
baseline <- tbod_schema(url, refresh = TRUE)
later <- tbod_check_schema(url, expected = baseline)
```

## Integrity and time budgets

Retrieval obtains a matching count and compares downloaded object IDs with the
applicable full or subset manifest. Missing, duplicate, or unexpected records
cause an error. For services with more than 10,000 matches and reliable
ordering and pagination, an incomplete finite preview of at most 1,000 rows
can use an ordered sample and verify its returned IDs with a subset manifest.
Other requests use the full matching ID manifest; `integrity = "full"`
requests that path explicitly. Provenance reports `subset-manifest`,
`full-manifest`, or `count-only` for `limit = 0` and never marks a deliberate
subset complete when more rows match. A full manifest of more than one
million matching records is rejected; narrow `where` or use an eligible
bounded preview.

These checks detect record-membership problems during a request, but a live
service can change record attributes between pages. A successful download is
not a frozen snapshot. The client does not keep a runtime cache of the data.

`timeout` limits each HTTP attempt. `total_timeout` sets an operation budget:
120 seconds for retrieval, 90 for catalog discovery, and 60 for metadata
inspection by default. Set `total_timeout = Inf` explicitly for an operation
without an overall deadline. Requests and retry waits use the remaining
budget. Parsing is checked after it finishes, so computation can exceed the
budget before a timeout is reported.

## Types and spatial results

Results preserve source column names and text values. ArcGIS
epoch-millisecond date fields become UTC `POSIXct` unless the source declares
an unknown time zone. In that case they remain raw milliseconds with a
warning; UTC editor-tracking creation and edit dates identified by layer
metadata are still converted. Date-only fields become `Date`; time-only and
timestamp-offset fields remain strings. Integer fields are converted where
safe, and large values retain an exact representation when possible.

Install the optional `sf` package for spatial retrieval. `spatial = TRUE`
returns an `sf` object; `out_sr = 4326` requests server projection to WGS 84.
Omit `out_sr` to keep the source's native CRS. Point, line, and polygon layers
are supported, including records with missing geometry. A nonspatial table
can be retrieved with `spatial = FALSE`.

```{r spatial}
projects <- tbod_get_dataset(
  "development-cases", spatial = TRUE, out_sr = 4326
)
boundary <- tbod_get_dataset(
  "city-boundary", fields = "OBJECTID", spatial = TRUE, out_sr = 4326
)
sf::st_crs(projects)$epsg
```

Both layers use one CRS so they can share a map. Keep only nonempty case
geometries for the display; the full `projects` object remains available for
analysis:

```{r spatial-map, fig.width=7, fig.height=5.5}
mapped_projects <- projects[!sf::st_is_empty(projects), ]

plot(
  sf::st_geometry(boundary),
  col = "#EDF4F6", border = "#7898A5", lwd = 1.2,
  axes = TRUE, graticule = TRUE, col_graticule = "#DCE7EB",
  xlab = "Longitude", ylab = "Latitude",
  main = "Active development cases in Tampa"
)
if (nrow(mapped_projects) > 0L) {
  case_fill <- grDevices::adjustcolor("#E76F51", alpha.f = 0.7)
  plot(sf::st_geometry(mapped_projects), add = TRUE, pch = 21,
       bg = case_fill, col = "#9B4934", cex = 0.85)
  graphics::legend(
    "bottomleft", legend = "Development case location", pch = 21,
    pt.bg = case_fill, col = "#9B4934", bg = "white",
    box.col = "#DCE7EB", cex = 0.85
  )
}
```

The checked `city-boundary` layer follows a shoreline representation and is
map context, not a legal survey boundary. See [coverage and source
limits](coverage.html) before interpreting a map or its record counts.

## Resume a large download

`tbod_download_dataset()` writes bounded RDS chunks to a local directory outside
the package source. This avoids collecting all feature pages in memory;
the full matching ID manifest still uses memory and retains the
one-million-match limit. Each chunk is a tibble or, with `spatial = TRUE`, an
`sf` object. `tbod_download_arcgis_layer()` offers the same path for a compatible
direct layer URL. For a stable ID from another ArcGIS Enterprise portal, pass
its sharing REST root through `portal`, as with `tbod_get_dataset()`.

```{r download}
saved <- tbod_download_dataset(
  "construction-permits",
  path = "~/tampaBayOpenData-downloads/permits",
  fields = c("OBJECTID", "RECORD_ID", "LASTUPDATE"),
  page_size = 500,
  total_timeout = 600
)
saved$complete
first_chunk <- readRDS(saved$files[[1]])
tbod_provenance(first_chunk)
```

Repeat the same call to resume. The client rechecks the live schema and
complete matching ID manifest, verifies saved chunks against checkpoints,
and downloads missing chunks. Changed query options, schema, CRS, endpoint,
or ID membership require a new directory. Temporary files from interrupted
writes are ignored. The directory must be empty on the first call, and
`resume = FALSE` does not overwrite an existing download. Chunk provenance
describes each chunk; `saved$complete` describes the whole matching download.
Record attributes may still change between requests or resumes, so the files
are not a frozen snapshot.

Downloads default to `total_timeout = Inf`; a finite deadline leaves
completed chunks available for a later resume. An exclusive directory lock
prevents concurrent writers. If an R process is terminated, confirm it has
stopped, then remove its empty `.lock` directory before resuming. Normal
completion and handled errors release the lock.

For source attribution and limitations, continue to [coverage and
provenance](coverage.html). For the catalog's scan behavior, see
[discovering data](discovery.html).
