---
title: "What makes up Tampa's mapped bicycle network?"
output: html_document
---

The City of Tampa calls this layer **Bike Lanes**, but its connection types also
include shared roadways, paths, trails, and other links. How does the mix change
when we count mapped segments versus sum the source's reported mileage?

This report retrieves the complete, current `bike-lanes` layer through
`tampaBayOpenData`. It requires an internet connection and the installed R
packages `tampaBayOpenData`, `knitr`, and `rmarkdown`. The source can change
between runs; the tables and charts below reflect the retrieval time printed
in the provenance table. No raw data file is bundled.

**Source:** [City of Tampa Bike Lanes layer and field definitions](https://arcgis.tampagov.net/arcgis/rest/services/OpenData/Transportation/MapServer/0)
and its [ArcGIS item and reuse information](https://tampa.maps.arcgis.com/home/item.html?id=d54aac9d4559488fa439e454df5f09ce).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.5)
library(tampaBayOpenData)
```

## Retrieve and verify the complete layer

`tbod_get_dataset()` checks the checked catalog schema and retrieves all
matching object IDs. Full retrieval compares the result with the service's
object-ID manifest. We stop if provenance reports an incomplete result or an
unexpected record count or duplicate ID.

```{r retrieve}
segments <- tbod_get_dataset(
  "bike-lanes",
  fields = c("OBJECTID", "CONN_TYPE", "MILEAGE", "ROADWAY"),
  page_size = 500,
  total_timeout = 120
)
source <- tbod_provenance(segments)
stopifnot(
  isTRUE(source$complete),
  identical(source$integrity, "full-manifest"),
  identical(source$validation_status, "checked"),
  nrow(segments) == source$matched_rows,
  nrow(segments) == source$returned_rows,
  !anyDuplicated(segments$OBJECTID)
)

knitr::kable(data.frame(
  publisher = source$publisher,
  retrieved_at_utc = format(source$retrieved_at, tz = "UTC", usetz = TRUE),
  matched_rows = source$matched_rows,
  returned_rows = source$returned_rows,
  complete = source$complete,
  integrity = source$integrity
))
knitr::kable(utils::head(segments, 5))
```

One record is one mapped line segment, not necessarily a whole route. The
publisher's `CONN_TYPE` coded-value domain supplies the labels below; this
report reads those labels from live layer metadata attached to the result.

```{r field-definitions}
connection_fields <- Filter(
  function(field) identical(field$name, "CONN_TYPE"), source$field_schema
)
stopifnot(length(connection_fields) == 1L)
codes <- connection_fields[[1]]$domain$codedValues
stopifnot(length(codes) > 0L)
code_names <- stats::setNames(
  vapply(codes, function(value) as.character(value$name), character(1)),
  vapply(codes, function(value) as.character(value$code), character(1))
)
knitr::kable(data.frame(
  code = names(code_names), publisher_label = unname(code_names)
))
```

## Check the measurements before adding them

`MILEAGE` is a publisher-supplied attribute; we do not remeasure geometry.
A zero may represent a short, rounded, or unmeasured connection, and its exact
meaning is not specified by the layer. Missing mileage remains in the segment
count but cannot contribute to a mileage sum. We also count missing class codes
and roadway labels so they cannot disappear from the story.

```{r quality}
connection_code <- as.character(segments$CONN_TYPE)
connection_code[is.na(connection_code) | !nzchar(trimws(connection_code))] <-
  "(missing)"
unknown_codes <- setdiff(unique(connection_code),
                         c(names(code_names), "(missing)"))
roadway <- trimws(as.character(segments$ROADWAY))
roadway[is.na(roadway)] <- ""

quality <- data.frame(
  check = c("Returned segments", "Missing connection code",
            "Codes absent from current metadata", "Missing reported mileage",
            "Zero reported mileage", "Negative reported mileage",
            "Blank or N/A roadway label"),
  value = c(nrow(segments), sum(connection_code == "(missing)"),
            length(unknown_codes), sum(is.na(segments$MILEAGE)),
            sum(segments$MILEAGE == 0, na.rm = TRUE),
            sum(segments$MILEAGE < 0, na.rm = TRUE),
            sum(toupper(roadway) %in% c("", "NA", "N/A")))
)
knitr::kable(quality)
if (length(unknown_codes)) print(unknown_codes)
if (any(segments$MILEAGE < 0, na.rm = TRUE)) {
  stop("Negative MILEAGE values need review before aggregation")
}

class_summary <- do.call(rbind, lapply(
  split(seq_len(nrow(segments)), connection_code),
  function(rows) {
    code <- connection_code[rows[1]]
    label <- unname(code_names[code])
    if (is.na(label)) label <- code
    data.frame(
      code = code, publisher_label = label,
      segments = length(rows),
      reported_miles = sum(segments$MILEAGE[rows], na.rm = TRUE),
      missing_mileage = sum(is.na(segments$MILEAGE[rows])),
      zero_mile_segments = sum(segments$MILEAGE[rows] == 0, na.rm = TRUE)
    )
  }
))
class_summary <- class_summary[order(-class_summary$reported_miles), ]
rownames(class_summary) <- NULL
knitr::kable(class_summary, digits = 2)
```

## Segment share versus reported-mile share

For a compact comparison, we group the City's codes into four families. These
family names are this report's analytical grouping; the preceding table keeps
the publisher's classes intact. Any new code remains in *Other or new code*.

```{r family-comparison, fig.height=5}
family_by_code <- c(
  SR_LT_LS = "Shared roadway", SR_MHT_LS = "Shared roadway",
  SR_LT_HS = "Shared roadway",
  BL_HT_HS = "Bike-lane categories", SBL_LT_LS = "Bike-lane categories",
  MUP = "Paths and trails", TR = "Paths and trails",
  DC = "Other connections", NG = "Other connections",
  SC = "Other connections"
)
family_name <- unname(family_by_code[connection_code])
family_name[is.na(family_name)] <- "Other or new code"
family <- do.call(rbind, lapply(split(seq_len(nrow(segments)), family_name),
  function(rows) data.frame(
    family = family_name[rows[1]], segments = length(rows),
    reported_miles = sum(segments$MILEAGE[rows], na.rm = TRUE)
  )
))
rownames(family) <- NULL
total_miles <- sum(segments$MILEAGE, na.rm = TRUE)
stopifnot(total_miles > 0)
family$segment_share_pct <- 100 * family$segments / nrow(segments)
family$mileage_share_pct <- 100 * family$reported_miles / total_miles
family_order <- c("Shared roadway", "Bike-lane categories",
                  "Paths and trails", "Other connections", "Other or new code")
family <- family[order(match(family$family, family_order)), ]
rownames(family) <- NULL
knitr::kable(family, digits = 2)

shares <- rbind("Segments" = family$segment_share_pct,
                "Reported miles" = family$mileage_share_pct)
old_par <- par(mar = c(4, 14, 3, 1))
barplot(shares, beside = TRUE, horiz = TRUE, names.arg = family$family,
        las = 1, col = c("#276fa6", "#e59a40"),
        xlim = c(0, max(shares) * 1.25),
        xlab = "Share of complete public layer (%)",
        main = "Mapped connection mix", legend.text = rownames(shares),
        args.legend = list(x = "topright", bty = "n"))
par(old_par)
largest_shift <- family[which.max(abs(family$segment_share_pct -
                                      family$mileage_share_pct)), ]
```

Counting segments gives each line one vote. Summing `MILEAGE` weights longer
lines more heavily. The two shares can therefore differ without implying a
data error. The largest gap in this retrieval is **`r largest_shift$family`**:
`r sprintf("%.1f", largest_shift$segment_share_pct)`% of segments versus
`r sprintf("%.1f", largest_shift$mileage_share_pct)`% of reported miles.

## Where are the zero-mileage records?

Segments with zero reported miles are still visible in the segment inventory.
The check below shows which publisher classes contribute these records.

```{r zero-mileage, fig.height=4.5}
zero_classes <- class_summary[class_summary$zero_mile_segments > 0, ]
if (nrow(zero_classes)) {
  zero_classes <- zero_classes[order(zero_classes$zero_mile_segments), ]
  old_par <- par(mar = c(4, 18, 3, 1))
  barplot(zero_classes$zero_mile_segments, horiz = TRUE,
          names.arg = zero_classes$publisher_label, las = 1,
          col = "#a45760", xlab = "Segments with zero reported MILEAGE",
          main = "Zero-mileage records by publisher class")
  par(old_par)
} else {
  cat("No zero-mileage segments in this run.\n")
}

cat(sprintf(
  "Retrieved %s mapped segments with %.2f summed reported miles at %s.\n",
  format(nrow(segments), big.mark = ","), total_miles,
  format(source$retrieved_at, tz = "UTC", usetz = TRUE)
))
```

## Interpretation and limits

The layer's name covers more than dedicated bike lanes. Neither a feature
count nor summed reported mileage establishes usable route length, current
condition, rider safety, or geographic access. The City's stress labels are
source classifications, not safety findings from this analysis. Overlapping
alignments may be counted more than once.

The package's complete-result check guards against missing or duplicate
record IDs and membership changes detected during retrieval. It cannot freeze
attribute values while the publisher edits the service. Review
`tbod_dataset_info("bike-lanes")$terms` and the linked publisher item before
reusing the data.
