---
title: "Explore Tampa bicycle network data"
output: rmarkdown::html_notebook
params:
  live: false
---

This notebook uses the `bike-lanes` entry in `tampaBayOpenData`. It renders
offline by default. To fetch a small live subset, set `live: true` above or run
`rmarkdown::render("bike-network.Rmd", params = list(live = TRUE))` from the
directory containing this file. Live chunks need an internet connection;
the map also needs the optional `sf` package.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Find the source

The checked catalog lookup does not contact the City.

```{r catalog}
library(tampaBayOpenData)

bike_catalog <- tbod_search_datasets("bike", source = "checked")
bike_catalog[, c("id", "title", "category", "verified")]

bike_info <- tbod_dataset_info("bike-lanes")
bike_info[c("title", "source_url", "verified", "scope_note")]
```

The `verified` date records a catalog endpoint check, not feature freshness.
Review `bike_info$terms` before using the data. The layer includes shared
roadways, paths, and trails as well as dedicated lanes.

## Retrieve and inspect 20 segments

This request uses exact source field names and orders by object ID so the
subset is reproducible while the source remains unchanged. The service is live,
so its records may change later.

```{r retrieve, eval=params$live}
bike <- tbod_get_dataset(
  "bike-lanes",
  fields = c("OBJECTID", "ROADWAY", "CONN_TYPE"),
  order_by = "OBJECTID ASC",
  limit = 20,
  page_size = 10
)
head(bike)
```

Provenance records the source, retrieval time, and whether the result covers
all matching records. A 20-record limit normally produces an incomplete subset.

```{r provenance, eval=params$live}
bike_source <- tbod_provenance(bike)
bike_source[c("source_url", "retrieved_at", "matched_rows", "returned_rows", "complete")]
```

The chart counts connection types **within the retrieved subset**. It is not
an estimate of the whole bicycle network.

```{r connection-types, eval=params$live}
connection_counts <- sort(table(bike$CONN_TYPE, useNA = "ifany"),
                          decreasing = TRUE)
connection_counts
if (length(connection_counts)) {
  barplot(connection_counts, las = 2, cex.names = 0.8,
          ylab = "Segments in retrieved subset",
          main = "Connection types in retrieved segments")
}
```

## Map the same subset (optional)

This makes a second live request. Its records can differ if the City updates
the layer between requests. `out_sr = 4326` asks the service for WGS 84
coordinates.

```{r map, eval=params$live}
if (requireNamespace("sf", quietly = TRUE)) {
  bike_map <- tbod_get_dataset(
    "bike-lanes",
    fields = c("OBJECTID", "ROADWAY", "CONN_TYPE"),
    order_by = "OBJECTID ASC",
    limit = 20,
    page_size = 10,
    spatial = TRUE,
    out_sr = 4326
  )
  sf::st_crs(bike_map)$epsg
  plot(sf::st_geometry(bike_map), main = "Retrieved bicycle network segments")
} else {
  message("Install sf to draw the map.")
}
```
